---
title: "Estero La Cruz Guild Analysis--Simplified"
author: "Pava Francisca Ellison Gonzalez"
date: "12/5/2020"
output:
  html_document: 
    fig_width: 8
    highlight: breezedark
    theme: cosmo
    toc: yes
    toc_float: yes
output_dir: "docs"
---

# Introduction

In this document, we'll summarize the final analysis of several years of waterbird surveys at Estero La Cruz in Bahía de Kino, Sonora. The point of all this is to address a simple question- Has the number of shorebirds at Estero La Cruz changed over time? 

```{r set up, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

# Los Datos

The first steps are loading all required packages and of course---the data. This may seem like a long list of packages, but I've only added in the important ones that I know are being applied. Any additional packages that those on this list are dependent upon will also be loaded. The list is pretty long so I'm tempted to put it at the end...but also thats bad coding I think

```{r pkgs, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2); theme_set(theme_linedraw())
library(ggpubr)
library(ggeffects)
library(gridExtra)
library(stringr)
library(lme4)
library(MASS)
library(stringr)
library(lubridate)
library(tidyr)
library(readr)
library(RColorBrewer)
library(MuMIn)
library(patchwork)
library(pscl)
library(StatisticalModels)
library(numDeriv)
library(multcomp)
library(AICcmodavg)
library(sjPlot)
```

All the necessary files should be saved in the working directory. 
The files we'll Counts of birds in each guild for each survey at Estero La Cruz from 2013/14 to present cycle and these are "environment variables" taken at each survey. Similar to the biodiversity 

```{r import data}
guilds <- read_csv("guilds_final.csv",
                         col_types = cols(Date = col_date(format = "%m/%d/%y",
                         na=0)))
```

Here's a sneak peak of the data--as you can see it's set up with each survey effort as a row with every guild as a column. 
`r kable(guilds[1:5, 1:5], caption = "Preview of Guild Totals")`


# First NBGLMER Modeling

***EXLPAIN AND DESCRIBE MODELS BETTEER TO EXPLAIN WHAT THE EFFECTS ARE***
random can be intercept or rate of change-
Here we'll use shorebirds as the example

I started playing with the idea of mixed models because they are a common tool used in biological modeling. Given the data we're wworking with and question we're trying to answer, the random effect that needs to be taken into account with this modeling is the observer(s). With several years of data and different people counting birds over the years, that's probably one of the biggest sources of variation in the data---and hence random effect of the observer(s)!! 

```{r poop, echo=FALSE}
options(na.action = "na.omit")
guilds$season <- as.character(guilds$szn)
guilds$ciclo <- guilds$temporada
```
Run the model:
```{r final shore nb glmer}
glmernb.shore1 <- glmer.nb(shore ~ temporada + season + (1|obs), data=guilds)
summary(glmernb.shore1)
GLMEROverdispersion(glmernb.shore1)
```
Alright- it looks like we aren't dealing with an absurd amount of overdispersion that was previously a problem. 

## Residuals

Now let's take a look at 'em. 
```{r nb glmer resids, fig.height=8, fig.width=8, echo=TRUE}
##put all the imporant things into a data frame
resid_shore1<-data.frame(
    pearsons_residuals=resid(glmernb.shore1,type="pearson"),
    deviance_residuals=resid(glmernb.shore1,type="deviance"),
    response_residuals=guild_totals$shore-predict(glmernb.shore1,type = "response"),
    predicted=predict(glmernb.shore1,type = "response"),
    observed=guild_totals$shore)

##plot predictions and resids
ggplot(resid_shore1,aes(x=predicted,y=pearsons_residuals))+
  geom_point() +
  ggplot(resid_shore1,aes(x=predicted,y=deviance_residuals))+
  geom_point()+
  ggplot(resid_shore1,aes(x=predicted,y=response_residuals))+
  geom_point()+
  ggplot(resid_shore1,aes(x=predicted,y=observed))+ geom_point()
```

Similar to previous models, this isn't necessarily the best looking set of residual graphs, but for now we can work with it.

## Combining Migration Seaons

I have only begun looking into the world of testing models and hypotheses. One method I encountered to test general hypotheses was an example I found on the internet proved useful. Essentially what I was interested in testing was whether or not there are differences between the seasons. This sets up a pairwise comparison between factors using a Tukey test. 

```{r hypothesis test}
posthoc1 <- glht(glmernb.shore1, linfct = mcp(season="Tukey"))
mcs <- summary(posthoc1, test=adjusted("single-step"))
mcs     ##call it to look at the pairwise comparison 
```

This is pretty neat. It shows which factors are siginificantly different from one another and which are similar enough to be the same. 
Since there are only 3 factors in the variable of season, the output is pretty small and easy to read. But imagine if there were 10 different factors of the variable? Could get pretty messy pretty quickly---so another way to look at each factor is to use letters. This is easier to follow and thus more accessible to interpret determine whether factors were similar or significantly different. 

```{r letters of hyp test}
cld(mcs, level=0.05, decreasing=TRUE)   ##cld = compact letter display
```

So what all that means is that spring and fall are similar to one another (hence the same letter) but winter is different. This makes sense given there is an influx of migrants in the winter and birds similarly moving through or completely absent in the fall and spring. In the OG data, 1=fall, 2=winter, 3=spring. Below I'll change the data so that fall/spring=1 and keep 2=winter using a simple `ifelse()` command. 

```{r combine seasons}
guilds$season <- ifelse(guilds$season != 2, 1, 2)
guilds$season <- as.factor(guilds$season)
```

Now that the data is combined, let's take a new look at it. Here I used more or less the same code as the boxplots at the beginning of the document. 

```{r combined szn plots, echo=FALSE}
shore1 <- ggboxplot(guilds, x = "ciclo", y = "shore", facet.by = "season",
          panel.labs = list(season=c("Otoño/Primavera", "Invierno")),
          ylab = "Shorebird Count", ggtheme = theme_linedraw())
shore1
```

And just a refresher/overview, let's look at the data, but rather than being divided by season, it is only divided by the work cycle/year ("ciclo").

```{r og data}
sb <- ggplot(guilds, aes(factor(ciclo), shore, group=temporada))
sb + geom_boxplot() + labs(title = "Shorebird Counts by Season") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Final Model Selection and Interpretation

Since I haven't been able to get the automated model selection `dredge()` to work with the negative binomial glmer, there are some other ways to do what I want to do and compare the models. It's not quite as pretty but as far as I understand model selection, it makes sense.
First we'll make a null model and go through some steps to change/add variables to our models.

```{r shorenb model final final}
null.shore <- glmer.nb(shore ~ 1 + (1|obs), data=guilds)

##and start with a simple model containing what we want to test
glmernb.shore1 <- glmer.nb(shore ~ ciclo + season + (1|obs), data=guilds)
summary(glmernb.shore1)
```

Just from an intial look at this model summary, it looks like there is a significant decline in the abundance of shorebirds over time. It also appears, as expected, that there is a significant increase in in the winter months when compared to the rest of the year. Before we saying anything more, let's do a handful of other steps to check the model and make sure this will be our final model.

```{r overdisp and anova}
GLMEROverdispersion(glmernb.shore1)
##now let's compare it with the null by making an ANOVA table
anova(null.shore, glmernb.shore1)
```

Since there is a significant difference, this means we can reject the null model and accept the simple negative binomial mixed model as a betterr fit to the data. I also feel OK about that given the AICc/BIC are slightly lower, and the log-Liklihood has moved a bit closer to 0 (which is another one of many things to look for). 

## Selecting a Model

There are a wide variety of criteria and methods that are used to select models. Here, I go through a few different 
Now let's add in some other potential variables that would affect the presence of birds, in this case, the tide!

```{r more models for comp}
##add tide height?
glmernb.shore2 <- glmer.nb(shore ~ ciclo + season + tide_height + (1|obs),
                           data=guilds)
##add tide dir?
glmernb.shore3 <- glmer.nb(shore ~ ciclo + season + tide_dir + (1|obs),
                           data=guilds)
##add both tide variables?
glmernb.shore4 <- glmer.nb(shore ~ ciclo + season + tide_height + tide_dir +
                             (1|obs), data=guilds)
```

We can make a neat model selection table based on information selection criteria. Here we will use the second order AIC (AICc) for comparison. For the model names, I abbreviated the names into "modX" so note that "mod1" = "glmernb.shore1", "mod2" = "glmernb.shore2", etc...

```{r aicc model table}
shore.cand <- list(null.shore, glmernb.shore1, glmernb.shore2, glmernb.shore3, glmernb.shore4)
mod.names <- list(c("null", "mod1", "mod2", "mod3","mod4"))
aictab(shore.cand, mod.names)       ##this exports our pretty html table below
```

Let's do another check by comparing increasingly complex models using ANOVAs. If the Chi-squared value is significant, that means that the model is a better fit. 
We will set up comparisons between our simplest model (glmernb.shore1) and each of the increasingly complex models. While we can use the code to set up single comparisons between models (ie. `anova(mod1, mod2)`) we can also combine it into one line of code such as `anova(mod1, mod2, mod3, ...)` and we will get a table that compares our simplest model (mod1) with the rest of the models. 

`r kable(anova(glmernb.shore1, glmernb.shore2, glmernb.shore3, glmernb.shore4))`

What each of these show, is that with adding in the height and/or direction of the tide, we do not get a better fitting model. That means we should stick to the first (and most simple!) model as is the best fit. 

Another potential approach to model selection would be to average the models. But rather than follow more coomplicated steps to do that, I think it's OK to stick with this process of selecting the best model

## Calculating R^2

Now that we have selected a model, one interesting piece of information about the model is checking the R^2 value. While this measure is commonly used in lms and glms it can be applied to mixed models. Here the function will calculate the R^2 values taking into account only the fixed effects (marginal) as well as all effects (conditionall) using 3 different methods.
Here's the explanataion taken from the `help("r.squaredGLMM")`: 

> Marginal R_GLMM² represents the variance explained by the fixed effects, and is defined as: $$R_GLMM(m)² = (σ_f²) / (σ_f² + σ_α² + σ_ε²)$$ 
Conditional R_GLMM² is interpreted as a variance explained by the entire model, including both fixed and random effects, and is calculated according to the equation: $$R_GLMM(c)² = (σ_f² + σ_α²) / (σ_f² + σ_α² + σ_ε²)$$ where σ_f² is the variance of the fixed effect components, σ_α² is the variance of the random effects, and σ_ε² is the “observation-level” variance.

So with that we can make a table of these values:
```{r r^2 shore mod, warning=FALSE}
(r.squaredGLMM(glmernb.shore1))
```

From the this, it looks like our model does a pretty decent job of fitting the data and explaining the variance. Given we're modeling change over time and not taking into account other environmental variables that are likely affecting the presence/absence/abundance of birds, I feel okay about believing this model. 

## Check Residuals

We'll run more or less the same code as we used above. See the .rmd file for code.
Recall that the difference now is that 

```{r final nb glmer resids, fig.height=8, fig.width=8, echo=FALSE}
##put all the imporant things into a data frame
resid_shore2<-data.frame(
    pearsons_residuals=resid(glmernb.shore1,type="pearson"),
    deviance_residuals=resid(glmernb.shore1,type="deviance"),
    response_residuals=guild_totals$shore-predict(glmernb.shore1,type = "response"),
    predicted=predict(glmernb.shore1,type = "response"),
    observed=guild_totals$shore)

##plot resids and all that jazz
pr <- ggplot(resid_shore2,aes(x=predicted,y=pearsons_residuals)) + geom_point()
dr <- ggplot(resid_shore2,aes(x=predicted,y=deviance_residuals)) + geom_point()
rr <- ggplot(resid_shore2,aes(x=predicted,y=response_residuals)) + geom_point()
po <- ggplot(resid_shore2,aes(x=predicted,y=observed)) + geom_point()
grid.arrange(pr, dr, rr, po)
```

Again, it looks more or less the same as before. Our residuals aren't perfectly randomly distributed but I'm not sure what I could do at this point to get a better fit. Given the process to get to this model, I think it's acceptable for the sake of this analysis. 

## Plotting Model Predictions

Now it's time to look at our model predictions. One of the most important aspects of a regression are the estimates of each variable in the model. Since we're working with negative binomial models, there is a log link meaning the effect estimates need to be log transformed. We can do that easily by using R as a fancy calculator. Before we do that, let's take a quick look at the different estimates of the model. 

```{r shore estimates graph, echo=FALSE}
tmp <- as.data.frame(confint(glht(glmernb.shore1))$confint)
tmp$Comparison <- rownames(tmp)
ggplot(tmp, aes(x = Comparison, y = Estimate, ymin = lwr, ymax = upr)) +
  geom_errorbar() + geom_point()
```

The first thing you might notice is that the intercept is very high- yes but it's not as important. If we think about the data we're modeling, there are counts of shorebirds in the thousands, and in the first year of data (2013-2014 season), there is a very high degree of variability.  
What we really want to pay more attention to are the estimates of two fixed effects of the model. Just by looking at this grarph, the effect of season is a positive increase while that of temporada(which is the year/cycle) is negative.  
If you recall from the summary output of the model the estimate of the variable temporada was: `r tmp$Estimate[2]`. To translate that into normal terms, as the year/cycle changes by 1, the change in shorebird abundance is multipled by a factor of `r exp(-0.21397)`. This can be interpreted into that "the abundance of shorebirds is declining by `r (1-exp(-0.21397))*100`% each year." 
One way we can visualize this is by plotting the predicted values on a graph. We can plot the overall trend of the model. Note that this will back-transform the predicted values to the original scale of our response variable.  

```{r main pred graph group, echo=FALSE}
pred1 <- ggpredict(glmernb.shore1, c("temporada"))
plot(pred1)
```

However, we know there are differences between the season and that there was a significant effect in the model. From the model estimate of that variable, we can say winter season sees an increase of `r exp(tmp$Estimate[3])*100`% in the number of shorebirds in comparison to the rest of the year. 
Knowing that the difference is so great, let's take a look at the same graph as above, but grouped by season. 
*Recall that the seasons are split into 1: Fall/Sping and 2: Winter.*

```{r shore pred graph, echo=FALSE}
pr1 <- ggpredict(glmernb.shore1, c("temporada", "season"))
plot(pr1, colors="ipsum")
##pick one
pd <- position_dodge(width=0.4) 
gg.s1 <- ggplot(pr1, aes(x, predicted, colour = group)) + scale_color_manual(values = c("#3f2d54", "#75b8d1"))
gg.s1 + geom_point(position = position_dodge(.3)) + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), position = pd) 
```
    
*Note that both graphs above are the same data, just plotted differently.*    
 
In both graphs, the seasons are split into 1 (Fall/Sping--the migration seasons) in purple and 2 (Winter) in blue. These colors will be consistent throughout the rest of this document. 
This shows that while shorebird abundance is declining in both seasons, it's clear that the decline is more pronounced in the winter-we can visually see this as the slope of the line is steeper. This could potentially be due a decline in the migratory species overwintering, so it would be a good idea to look further into the migratory species.

Plot the random effects. 
And also a diagnostic plot:
> For **generalized linear mixed models**, returns the QQ-plot for random effects.   

So diagnostic plot is essentially just a qq plot

```{r more}
set_theme(base = theme_linedraw())
plot_model(glmernb.shore1, type="re", colors = "ipsum", vline.color = "red")
plot_model(glmernb.shore1, type="diag")
```

The first plot shows the random effects. The red line indicates no effect. The blue indicates a positive effect while purple is negtative. 
The second plot above, as mentioned, is a qqplot. 

```{r más}
##show it as connected line
shore_resids <- cbind(resid_shore2, guild_totals$season)
shore_resids <- cbind(shore_resids, guild_totals$temporada)
shore_resids$temporada <- shore_resids$`guild_totals$temporada`
shore_resids$season <- shore_resids$`guild_totals$season`
shore_resids$szn <- as.factor(shore_resids$`guild_totals$season`)

ggplot(shore_resids, aes(guild_totals$temporada, predicted, colour = szn)) + 
  stat_summary(geom="smooth", fun.data=mean_cl_normal, width=0.1, conf.int=0.95) +
  stat_summary(geom="line", fun=mean, linetype="dashed") +
  stat_summary(geom="point", fun=mean) 
```


# Convergence Problem Troubleshooting

Both during my model exploration with shorebirds as well as with other guilds, I ran into a handful of problems. One common problem I ran into was a failure for the model to converge. This convergence failure essentially means that the model fits the data extremely poorly, or the observations just don't match up. But this isn't the end of the world. For example, when I tried to apply a negative binomial mixed model to the waterfowl guild data, I ran into this problem. Below is an example of how I handled it.

```{r poop 3, echo=FALSE}
options(na.action = "na.omit")
```

```{r fowl}
##now the other guilds. #this outputs a lot of warnings and it isn't goood 
glmernb.fowl1 <- glmer.nb(waterfowl ~ temporada + season + (1|obs), data=guild_totals)
summary(glmernb.fowl1)
```

Luckily, I found this [neat guide](https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html) that I followed and found helpful. Below I follow the steps of the aformentioned guide and fix the problem. What the steps below basically do is try to optimize the model, starting from the inital fit but increasing the number of iterations. 

```{r convergence problems}
tt <- getME(glmernb.fowl1,"theta")
ll <- getME(glmernb.fowl1,"lower")
min(tt[ll==0])

derivs1 <- glmernb.fowl1@optinfo$derivs
sc_grad1 <- with(derivs1, solve(Hessian,gradient))
max(abs(sc_grad1))

max(pmin(abs(sc_grad1),abs(derivs1$gradient)))

ss <- getME(glmernb.fowl1,c("theta","fixef"))
glmernb.fowl2 <- update(glmernb.fowl1,start=ss,control=glmerControl(optCtrl=list(maxfun=2e4)))
```

No more failure to converge! So let's check for overdispersion and look at the summary output of the model.

```{r check model}
GLMEROverdispersion(glmernb.fowl2)
summary(glmernb.fowl2)
```
It looks like there is a significcant decline in the abundance of waterfowl over time. It also appears, as expected, that there is a significant increase in in the winter months in comparison to the rest of the year.  

# Final Results of Other Guilds

Below, I present mostly just the numbers from the other guilds. 

## Waterfowl
Since we just worked on the waterfowl model above, let's start there and begin by plottin the data. 
```{r plot fowl, echo=FALSE}
water1 <- ggboxplot(guild_totals, x = "temporada", y = "waterfowl", 
                   ylab = "Count", ggtheme = theme_linedraw())
water1
#labs(title="Count grouped by season") + theme(plot.title = element_text(hjust = 0.5))
water2 <- ggboxplot(guild_totals, x = "temporada", y = "waterfowl", facet.by = "season", 
                   panel.labs = list(season=c("Otoño/Primavera", "Invierno")), 
                   ylab = "Count", ggtheme = theme_linedraw())
water2
#labs(title="Count Divided by season") + theme(plot.title = element_text(hjust = 0.5))
```

Recall our model `glmernb.fowl2` from directly above.  
Similar to before, we can once again calculate the R^2 values. This time we'll show it in neat table. `r kable(r.squaredGLMM(glmernb.fowl2))`
Once again, it's looking like we've got a pretty decent looking model.  

Let's check the residuals
```{r fowl nb glmer resids, fig.height=8, fig.width=8, echo=FALSE}
##put all the imporant things into a data frame
resid_fowl<-data.frame(
    pearsons_residuals=resid(glmernb.fowl2,type="pearson"),
    deviance_residuals=resid(glmernb.fowl2,type="deviance"),
    response_residuals=guild_totals$waterfowl-predict(glmernb.fowl2,type = "response"),
    predicted=predict(glmernb.fowl2,type = "response"),
    observed=guild_totals$waterfowl)

##plot resids and all that jazz
pr.f <- ggplot(resid_fowl,aes(x=predicted,y=pearsons_residuals)) + geom_point()
dr.f <- ggplot(resid_fowl,aes(x=predicted,y=deviance_residuals)) + geom_point()
rr.f <- ggplot(resid_fowl,aes(x=predicted,y=response_residuals)) + geom_point()
po.f <- ggplot(resid_fowl,aes(x=predicted,y=observed)) + geom_point()
grid.arrange(pr.f, dr.f, rr.f, po.f)
```

And now let's look at the model predctions. Using the estimate of our model, we can say there has been a decline of `r (1-exp(-0.36894))*100`% over the seasons and an increase of `r (exp(1.29668)*100)`% in the winter. 

```{r fowl mod plots, echo=FALSE}
pr.fowl <- ggpredict(glmernb.fowl2, c("temporada", "season"))
plot(pr.fowl, colors="ipsum")

gg.f1 <- ggplot(pr.fowl, aes(x, predicted, color=group)) + scale_color_manual(values = c("#3f2d54", "#75b8d1"))
gg.f1 + geom_point(position=pd) + geom_errorbar(aes(ymin=conf.low,ymax=conf.high), 
                                                 position=pd) + 
  scale_x_continuous(breaks = 14:20) + labs(x = "Year", y= "Predicted Count")
```

Similar to the shorebirds, we can see a similar difference in that the decline seems much more stark in the winter.  

## Gulls/Terns/Skimmers

Check the data
```{r gull plot, echo=FALSE}
gulls.terns1 <- ggboxplot(guild_totals, x = "temporada", y = "gulls_terns",
                ylab = "Count", ggtheme = theme_linedraw())
gulls.terns1

gulls.terns2 <- ggboxplot(guild_totals, x = "temporada", y = "gulls_terns", 
                          facet.by = "season", 
                          panel.labs = list(season=c("Otoño/Primavera", "Invierno")),
                          ylab = "Count", ggtheme = theme_linedraw())
gulls.terns2
```

Make the model
```{r gulls/terns/allies}
glmernb.gulls <- glmer.nb(gulls_terns ~ temporada + season + (1|obs),
                          data=guild_totals)
summary(glmernb.gulls)
GLMEROverdispersion(glmernb.gulls)
```
Once again, it looks likee there is a significant decline in the abundance of gulls/terems/skimmers over time as well as a significant increease in the winter due to the influx of migrants. 
From the model estimates, there is a decline of `r (1-exp(-0.04662))*100`% of the in gulls/terns/skimmers guild. While this isn't as exaggerated as the other guilds, it is still something worth noting.  

Calculate R^2 values
`r kable(r.squaredGLMM(glmernb.gulls))`

Review the residuals
```{r gull glmer resids, fig.height=8, fig.width=8, echo=FALSE}
##put all the imporant things into a data frame
resid_gulls<-data.frame(
    pearsons_residuals=resid(glmernb.gulls,type="pearson"),
    deviance_residuals=resid(glmernb.gulls,type="deviance"),
    response_residuals=guild_totals$gulls_terns-predict(glmernb.gulls,type = "response"),
    predicted=predict(glmernb.gulls,type = "response"),
    observed=guild_totals$gulls_terns)

##pgot resids and all that jazz
pr.g <- ggplot(resid_gulls,aes(x=predicted,y=pearsons_residuals)) + geom_point()
dr.g <- ggplot(resid_gulls,aes(x=predicted,y=deviance_residuals)) + geom_point()
rr.g <- ggplot(resid_gulls,aes(x=predicted,y=response_residuals)) + geom_point()
po.g <- ggplot(resid_gulls,aes(x=predicted,y=observed)) + geom_point()
grid.arrange(pr.g, dr.g, rr.g, po.g)
```

Plot the model predictions  
```{r gull mod plots, echo=FALSE}
pr.gull <- ggpredict(glmernb.gulls, c("temporada", "season"))
gg.g1 <- ggplot(pr.gull, aes(x, predicted, color=group)) + scale_color_manual(values = c("#3f2d54", "#75b8d1"))
gg.g1 + geom_point(position=pd) + geom_errorbar(aes(ymin=conf.low,ymax=conf.high), 
                                                 position=pd) + 
  scale_x_continuous(breaks = 14:20) + labs(x = "Year", y= "Predicted Count") 
```

Here we can visualize that the decline is not as steep as the other guilds, but still present. We can also once again see the notable differences between the winter and non-winter seasons. 

## Pelicans/Cormorants/Allies

Plot the data
```{r plot pel, echo=FALSE}
pel.corm1 <- ggboxplot(guild_totals, x = "temporada", y = "pel_corm",  
                   ylab = "Count", ggtheme = theme_linedraw())
pel.corm1
pel.corm2 <- ggboxplot(guild_totals, x = "temporada", y = "pel_corm", facet.by = "season", 
                   panel.labs = list(season=c("Otoño/Primavera", "Invierno")), 
                   ylab = "Count", ggtheme = theme_linedraw())
pel.corm2
```

Making the model:
```{r pel corm mod}
glmernb.pel <- glmer.nb(pel_corm ~ temporada + season + (1|obs),
                          data=guild_totals)
ss <- getME(glmernb.pel,c("theta","fixef"))
pelmod <- update(glmernb.pel,start=ss,control=glmerControl(optCtrl=list(maxfun=2e4)))
summary(pelmod)
GLMEROverdispersion(pelmod)
```
Pelicans/cormorants/allies guild is declining by `r (1-exp(-0.22764))*100`% each year. 

Calculate R^2 values
`r kable(r.squaredGLMM(pelmod))`

Plot model predictions
```{r pel mod plots, echo=FALSE}
pr.pel <- ggpredict(pelmod, c("temporada", "season"))
gg.p1 <- ggplot(pr.pel, aes(x, predicted, color=group)) + scale_color_manual(values = c("#3f2d54", "#75b8d1"))
gg.p1 + geom_point(position=pd) + geom_errorbar(aes(ymin=conf.low,ymax=conf.high), 
                                                 position=pd) + 
  scale_x_continuous(breaks = 14:20) + labs(x = "Year", y= "Predicted Count")
```
Something to note here is the axis. If you look at the boxplots above, you'll see there are two very high counts (ploted as dots aka outliers) in the 2013-2014 season. Now that can very obviously be skewing the data, but take a second to look at the axis of our graphed model predictions and you'll see that it's much smaller. 

## Long-legged Waders

**The only guild that lacks a final model**

```{r plot waderz, echo=FALSE}
waders1 <- ggboxplot(guild_totals, x = "temporada", y = "waders",
                  ylab = "Count", ggtheme = theme_linedraw())
waders1

waders2 <- ggboxplot(guild_totals, x = "temporada", y = "waders", facet.by = "season", 
                   panel.labs = list(season=c("Otoño/Primavera", "Invierno")), 
                  ylab = "Count", ggtheme = theme_linedraw())
waders2
```

```{r waders mod}
glmernb.waders <- glmer.nb(waders ~ temporada + season + (1|obs),
                          data=guild_totals)
summary(glmernb.waders)
GLMEROverdispersion(glmernb.waders)
```

***As you can see, this model failed to onverge and I haven't worked on getting a better fit yet***

## Final Table of All Models

We can put a summary of all the important parts of each model into a single table. Note that here the "estimates" of each variable have been back transformed. For example, in the shorebird model, rather than display the estimate of "-0.21397" for temporada (aka the year), the table will show the back transformed intercept (recall that for a negative binomial model we have to use the `exp()` function) so the number displayed is the incidence rate ratio of `r exp(-0.21397)`. 
This simply means that as that variable increases by 1 unit, the response variable is multipled by a factor of `r exp(-0.21397)`. 
Also note that the R^2 values displayed are calculated using the log-normal method that was shown in the tables above. 

```{r guild mods table}
tab_model(glmernb.shore1, glmernb.fowl2, glmernb.gulls, pelmod, 
          dv.labels = c("Shorebirds", "Waterfowl", "Gulls/Terns/Skimmers", 
                        "Pelicans/Cormorants/Allies"),
          string.ci = "Conf. Int (95%)", 
          string.p = "P-Value", pred.labels = c("Intercept", "Year", "Season"))
```


# Migrants and Residents

Another important grouping to make in the bird is between the migrantory and resident species for many reasons. One interesting reason just based off the guild muilds is that when looking at the plotted model predictions, it seemed that in general, the decrease in birds over time was more pronounced in the winter season.   

```{r import mig data}
mig <- read_csv("mig_final.csv", col_types = cols(Date = col_date(format = "%m/%d/%y")))
mig <- filter(mig, szn != "4")               ##remove summer
mig$season <- ifelse(mig$szn != 2, 1, 2)     ##combine spring/fall
mig$ciclo <- mig$temporada                   ##just to give it a diff name
colnames(mig)     ##check that data combined and is complete
```

## Migrants

Start with migrants. 
Plot data:
```{r plot mig, echo=FALSE}
ggboxplot(mig, x = "ciclo", y = "migrants", ylab = "# Individuals")

migrants <- ggboxplot(mig, x = "ciclo", y = "migrants", facet.by = "season",
          panel.labs = list(season=c("Otoño/Primavera", "Invierno")))
migrants
```

Make the model
```{r model}
mig.mod1 <- glmer.nb(migrants ~ ciclo + season + (1|obs), data = mig)
summary(mig.mod1)
```
`r (1-exp(-0.13459))*100`% decrease over the years
`r exp(0.78086)*100`% higher in the winter
 
Plot the residuals
```{r mig glmer resids, fig.height=8, fig.width=8, echo=FALSE}
##put all the imporant things into a data frame
resid_mig<-data.frame(
    pearsons_residuals=resid(mig.mod1,type="pearson"),
    deviance_residuals=resid(mig.mod1,type="deviance"),
    response_residuals=guild_totals$gulls_terns-predict(mig.mod1,type = "response"),
    predicted=predict(mig.mod1,type = "response"),
    observed=mig$migrants)

##pgot resids and all that jazz
pr.m <- ggplot(resid_mig,aes(x=predicted,y=pearsons_residuals)) + geom_point()
dr.m <- ggplot(resid_mig,aes(x=predicted,y=deviance_residuals)) + geom_point()
rr.m <- ggplot(resid_mig,aes(x=predicted,y=response_residuals)) + geom_point()
po.m <- ggplot(resid_mig,aes(x=predicted,y=observed)) + geom_point()
grid.arrange(pr.m, dr.m, rr.m, po.m)
```

Calculate R^2 values
`r kable(r.squaredGLMM(mig.mod1))`

Plot model predictions

First we'll look at the effects of year on the model, with the other effects considereed constants. 
```{r mig mod plot, echo=FALSE}
pr.mig <- ggpredict(mig.mod1, c("ciclo"))
plot(pr.mig, color="ipsum")
```

And now compare that graph to the model, but dividing out the season as we 
```{r mig mod plot pt 2, echo=FALSE}
pr.mig1 <- ggpredict(mig.mod1, c("ciclo", "season"))
plot(pr.mig1, color="ipsum")       ##smoothed line

##turn into gg obj and then point/errorbar, as done with guild prediction graphs
gg.mig1 <- ggplot(pr.mig1, aes(x, predicted, color=group)) + scale_color_manual(values = c("#3f2d54", "#75b8d1"))
gg.mig1 + geom_point(position=pd) + geom_errorbar(aes(ymin=conf.low,ymax=conf.high), 
                                                 position=pd) + 
  scale_x_continuous(breaks = 14:20) + labs(x = "Year", y= "Predicted Count")
```

## Residents
*Repeat workflow*
Plot data: 
```{r plot res, echo=FALSE}
ggboxplot(mig, x = "ciclo", y = "residents", ylab = "# Individuals")
residents <- ggboxplot(mig, x = "ciclo", y = "residents", facet.by = "season",
                      panel.labs = list(season=c("Otoño/Primavera", "Invierno")),
                      title = "Residents")
residents
```

Make the model

```{r res model}
res.mod1 <- glmer.nb(residents ~ ciclo + season + (1|obs), data = mig)
summary(res.mod1)
```

Residents, although decreasing it's not significant. While they aren't necessarily increasing either, this means that the birds are probably fluctuating year to year but overall are not changing. This is good!!
Also, similar to all other groups, the numbers of residents increases `r exp(0.82429)*100`% in the winter months. 

Plot the residuals
```{r res glmer resids, fig.height=8, fig.width=8, echo=FALSE}
##put all the imporant things into a data frame
resid_res<-data.frame(
    pearsons_residuals=resid(res.mod1,type="pearson"),
    deviance_residuals=resid(res.mod1,type="deviance"),
    response_residuals=guild_totals$gulls_terns-predict(res.mod1,type = "response"),
    predicted=predict(res.mod1,type = "response"),
    observed=mig$residents)

##pgot resids and all that jazz
pr.r <- ggplot(resid_res,aes(x=predicted,y=pearsons_residuals)) + geom_point()
dr.r <- ggplot(resid_res,aes(x=predicted,y=deviance_residuals)) + geom_point()
rr.r <- ggplot(resid_res,aes(x=predicted,y=response_residuals)) + geom_point()
po.r <- ggplot(resid_res,aes(x=predicted,y=observed)) + geom_point()
grid.arrange(pr.r, dr.r, rr.r, po.r)
```

Calculate R^2 values
`r kable(r.squaredGLMM(res.mod1))`

Plot model predictions
```{r res mod plots, echo=FALSE}
pr.res <- ggpredict(res.mod1, c("ciclo", "season"))
plot(pr.res, color="ipsum")
gg.res1 <- ggplot(pr.res, aes(x, predicted, color=group)) + scale_color_manual(values = c("#3f2d54", "#75b8d1"))
gg.res1 + geom_point(position=pd) + geom_errorbar(aes(ymin=conf.low,ymax=conf.high), 
                                                 position=pd) + 
  scale_x_continuous(breaks = 14:20) + labs(x = "Year", y= "Predicted Count") 
```
As we can see in the model prediction graphs above- our confidence intervals are very large. And though the trend is a decrease over time, there is not a lot to support that statement. So in looking at these graphs, we can see how much overlap there is in the points and confidence intervals.


## Final Table of Mig Models

Repeating what we did, we can create a model summary table for our two. And once again, what were the intercepts of the model have been back transformed into incidence rate ratios. 

```{r mig/res mods table}
tab_model(mig.mod1, res.mod1, 
          dv.labels = c("Migrants", "Residents"),
          string.ci = "Conf. Int (95%)", 
          string.p = "P-Value", pred.labels = c("Intercept", "Year", "Season"))
```

This is a clean way to summarize some stuff.

## Some Thoughts

Given that most of the guilds seem to be trending downward. Let's take a look once again at the predicted values of the models for each guild. In each of the models, the starting point was much higher in the winter and seemingly decreased more notably.

```{r all plots together, echo=FALSE}
gr1 <- ggplotGrob(plot(pr1, show.title=FALSE, show.legend=TRUE, color="ipsum"))
gr2 <- ggplotGrob(plot(pr.fowl, show.title=FALSE, show.legend=FALSE, color="ipsum"))
gr3 <- ggplotGrob(plot(pr.gull, show.title=FALSE, show.legend=TRUE, color="ipsum"))
gr4 <- ggplotGrob(plot(pr.pel, show.title=FALSE, show.legend=FALSE, color="ipsum"))

grid.arrange(gr1, gr2, gr3, gr4, 
             layout_matrix = rbind(c(1,1,1,1,1,1,2,2,2,2,2),
                                   c(3,3,3,3,3,3,4,4,4,4,4)))
```

One reason I had moved from working with biodiversity indices to  working on this modeling project was because it was clear that important aspect of the survey data was missing. For the calculating the biodiversity indices for example, birds had to be identified to species. This completely glosses over the impact of larger aggregations of birds that are identified to a group but not necessarily species (ie estimations of "1000 peeps"). In separting the birds into guilds, this important aspect of data that can be captured. 

```{r mig res mod plots, fig.height=8, fig.width=7}
gm1 <- ggplotGrob(plot(pr.mig1, show.title=FALSE, show.legend=TRUE, color="ipsum"))
gre1 <- ggplotGrob(plot(pr.res, show.title=FALSE, show.legend=TRUE, color="ipsum"))
grid.arrange(gm1, gre1)
```

So in looking at our model, it doesn't seem as if residents are changing but migrants are. This might suggest that the overall declines across the bird guilds are unequally attributed to declines in migrants overwintering. This could be a big jump to make, but it's something that seemingly makes sense. 
BUT ALSO with the separation between migrant/resident spp, I did not include these unidentified birds. Despite this, we still saw some interesting changes and that migrants are following these trends to be declinging. 

But finally, a big thing to keep in mind is that there is a lot of variability in all our data over time **so all results should be taken with a grain of salt.**

# Future Directions

While this was fun, the work isn't done. For example, there are several other cool questions that can be answered. Part of the problem is the surveys lack a consistent protocol. For example, I do now know how people in 2015 measured the tide or temperature. Gathering better data on historic weather/climate conditions would be interesting to look at changes or potential correlations between temp and bird relative abundances. Though we take data on the temperature during each monitoring effort, examining climate from a wider lens would be interesting. Precipitation patterns or extreme temperature events could potentially be affecting the phenology of migration in some birds that may be using temp/precip patterns as environmental cues. While some temperature and precipitation data are available [here](http://clicom-mex.cicese.mx/mapa.html), I found the data in the area to be incomplete and lacking the last few years. 
Another interesting direction would be to relate the abundances of shorebirds (and/or other guilds) to tides. While we take data on the tides, it would be possible to more accurately characterize the tide.
